---
title: "Evironmental Injustice in the Inland Empire, CA"
author: "Kylie Newcomer"
format: html
warning: false
error: false
message: false
---

```{r}
library(tidyverse)
library(sf)
library(here)
library(stars)
library(tmap)

# read in geodatabase of EJScreen data at the Census Block Group level
ejscreen <- sf::st_read(here::here("data", "ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")) 

# filter to a state you are interested in
california <- ejscreen %>%
  dplyr::filter(ST_ABBREV == "CA") 

# filter to a county you are interested in
# Selecting counties in the Inland Empire
ie <- ejscreen %>%
  dplyr::filter(CNTY_NAME %in% c("Riverside County", "San Bernardino County"))



# find the average values for all variables within counties
california_counties <- aggregate(california, by = list(california$CNTY_NAME), FUN = mean)
```


### Inital Data Exploration

```{r}
tm_shape(ie) +
  tm_graticules() +
  tm_polygons(fill = "LOWINCPCT", 
              fill.legend = tm_legend(title = "Percent of Low \nIncome Households", 
                                      orientation = "portrait", 
                                      tm_pos_out = c("center", "right"),
                                      frame = FALSE),
              fill.scale = tm_scale_continuous(values = "-magma", 
                                               labels = c("0%", "20%", "40%", 
                                                          "60%", "80%", "100%"))) +
  tm_scalebar(position = tm_pos_in("right", "top")) +
  tm_compass(position = c("left", "top")) +
  tm_title("Distribution of Low Income Households in the Inland Empire")

```

```{r}
tm_shape(ie) +
  tm_graticules() +
  tm_polygons(fill = "PRE1960PCT", 
              fill.legend = tm_legend(title = "Percent of Homes \nwith Lead Paint", 
                                      tm_pos_out = c("center", "bottom"),
                                      orientation = "portrait",
                                      frame = FALSE), 
              fill.scale = tm_scale_continuous(values = "-magma", 
                                               labels = c("0%", "20%", "40%", 
                                                          "60%", "80%", "100%"))) +
  tm_scalebar(position = tm_pos_in("right", "top")) +
  tm_compass(position = c("left", "top")) +
  tm_title("Distribution of Homes with Lead Paint in the Inland Empire")
  
```

### Creating my variables of interest

```{r}
# Looking at children in poverty
ie_kids <- ie %>% 
  select(UNDER5, LOWINCPCT) %>%
  mutate(kid_inc = (UNDER5*LOWINCPCT)) # Number of children multiplied by percent in poverty

# Checking the min and max to see the range
min(ie_kids$kid_inc)
max(ie_kids$kid_inc)
```


```{r}
# Want to see how that compares to children living in homes with lead pain
ie_lead <- ie %>% 
  select(UNDER5, PRE1960PCT) %>% 
  mutate(kid_lead = (UNDER5*PRE1960PCT)) # Number of children multiplied by percent of homes built pre 1960

# Checking the min and max to see the range
min(ie_lead$kid_lead)
max(ie_lead$kid_lead)
```

## Environmental Injustice Maps

```{r}
tm_shape(ie_kids) +
  tm_graticules() +
  tm_polygons(fill = "kid_inc", 
              fill.legend = 
                tm_legend(title = "Estimated Number of Children", 
                                      tm_pos_out = c("center", "bottom"),
                                      orientation = "portrait",
                                      frame = FALSE), 
              fill.scale = tm_scale_intervals(values = "-magma", 
                                              n = 8)) +
  tm_scalebar(position = tm_pos_in("right", "top")) +
  tm_compass(position = c("left", "top")) +
  tm_title("Children Living in Poverty in the Inland Empire")
```

```{r}
tm_shape(ie_lead) +
  tm_graticules() +
  tm_polygons(fill = "kid_lead", 
              fill.legend = 
                tm_legend(title = "Estimated Number of Children", 
                                      tm_pos_out = c("center", "bottom"),
                                      orientation = "portrait",
                                      frame = FALSE), 
              fill.scale = tm_scale_intervals(values = "-magma")) +
  tm_scalebar(position = tm_pos_in("right", "top")) +
  tm_compass(position = c("left", "top")) +
  tm_title("Children Living in Homes with Lead Paint in the Inland Empire")
```

The data contains population demographic and environmental factors throughout the US. I focused on the Inland Empire (Riverside and San Bernardino county) of California, because of the fast growing communities and relatively "affordable" living compared to other counties in Southern California. I started by mapping my factors of interest (poverty levels and potential homes with lead paint). I wanted to specifically focus on quality of life for children under 5 years old, starting with estimating how many children are living in poverty. After seeing regions with a large amount children living in poverty, I compared that to how many children were living in homes with lead paint. There did not seem to be a correlation between regions with high poverty levels and high percentages of homes with lead paint. I attempted other visualizations to see if there were increased health or environmental threats to the children in areas with high poverty levels but was unable to come to any conclusions.

